Mapping

### 2025-04-10 episodes/snps-phylogeny.Rmd
Finished SNPs phylo

:::::::::::::
Snippy is fast but a single run still takes about 15 minutes. We would therefore tell snippy to run all the samples after each other. However this time we cannot use a wildcard to do so. We would instead run all the samples in a loop. The "real" loop is commented out, and we use one that runs only the sample we've analyzed so far.

@@

Now copy a reduced version of the results on the other samples, save it in the `results` folder. We will also rename the existing `snps` folder to save it.

@@

The folders don't contain everything that is output by `snippy`, to save space. But it is enough to run `snippy-core`.

@@

```bash
cat /proj/g2020004/nobackup/3MK013/<username>/molepi/results/snps/ERR029207/snps.txt
```

@@

@@

@@

Have a look at the content of these three files with `cat` or `head`.

@@

What is the difference between these files? Why is core.aln smaller than ?
What is in core.aln?

@@

Phylogenetic trees have been discussed during the lectures. We will here establish a phylogenetic tree from the file 'core.aln' with [IQ-TREE](http://www.iqtree.org/). IQ-TREE is a phylogeny software based on the maximum-likelihood principle. IQ-TREE has been widely used and cited. There is a very wide range of parameters that need to be chosen, such as nucleotide or amino-acid substitution models.

In our case, we want IQ-TREE to automatically select the best substitution model. IQ-TREE does that by testing many (among a very large collection) substitution models. We also have SNP data, which by definition do not contain constant (invariable) sites. We thus input the alignment with the `-s` option and the model with `-m MFP+ASC`. `MFP` will tell IQ-TREE to test a range of models, and `ASC` will correct for the fact that there is no constant sites.

```bash
module load iqtree
iqtree2 -s core.aln -m MFP+ASC
```

@@

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
cd tree
mv core.aln.treefile core.aln.newick 
cat core.aln.newick
```

@@

Hint: The log file of IQ-TREE contains a lot of information.

@@

IQ-TREE chose the TVMe+ASC model. We've discussed the ASC above, and you can read more about the TVMe in the [IQ-TREE manual](https://www.iqtree.org/doc/Substitution-Models). In short, it is a mdoel where base frequencies are deemed equal, and all rates estimated independently, except the rates of transversions (A <-> G and C <-> T) are equal.

@@

@@

>>>>

:::::::::::::

Snippy is fast but a single run still takes about 15 minutes. We would therefore tell snippy to run all the samples after each other. However this time we cannot use a wildcard to do so. We would instead run all the samples in a loop. The "real" loop is commented out, and we use one that runs only the sample we've analyzed so far.

@@

Now copy a reduced version of the results obtained from the other samples, save it in the `results` folder. We will also rename the existing `snps` folder to save it. We extract them from the archive in the `data` folder, using `tar`. `tar` is a program to take multiple files, including their folder structure and put them in a single `tar` "ball". Often the tarballs are compressed with `gzip`.

@@

Note the unusual way to pass arguments to `tar` with just `xvzf` instead of `-x -v -z -f` or `-xvzf`. `x` tells tar to extract the archive (`c` would be used to create the archive); `v` is for verbose, providing more information on the output; `z` tells tar to decompress the file; `f` tells what file to work on.

The folders don't contain everything that is output by `snippy`, to save space. But it is enough to run `snippy-core`, which summarizes several `snippy` runs.

@@

From the `molepi/results/snps` folder:

```bash
cat ERR029207/snps.txt
```

@@

```bash
for accession in ERR*; do
  echo <accession>
  cat <accession/snps.txt | grep <pattern>
done
```

:::::::::::::::  instructor

```bash
for accession in ERR*; do
  echo "${accession}"
  cat "${accession}"/snps.txt | grep "Variant-SNP"
done
```

@@

Instead of writing all samples explicitly, we could use wildcards:

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results/snps
snippy-core --ref=../../data/GCF_000195955.2_ASM19595v2_genomic.fna ERR*
```

@@

Have a look at the content of these three files>

@@

What is the difference between these files? Why is `core.aln` smaller than `core.full.aln`?
What is in `core.aln`?

@@

We will here infer a phylogenetic tree from the file 'core.aln' with [IQ-TREE](http://www.iqtree.org/).

In this case, we want IQ-TREE to automatically select the best (nucleotide) substitution model. IQ-TREE does that by testing many (among a very large collection) substitution models. We also have SNP data, which by definition do not contain constant (invariable) sites. We thus input the alignment with the `-s` option and the model with `-m MFP+ASC`. `MFP` will tell IQ-TREE to test a range of models, and `ASC` will correct for the fact that there is no constant sites.

Infer a phylogenetic tree with IQ-TREE, using the aligned polymorphic sites identified by `snippy-core` above. IQ-TREE should find the best possible model, adding the correction for no constant sites. Don't forget to load the right module.

::: hint

```bash
iqtree -h
```

::::::::

```bash
module load <module>
iqtree2 <alignment> <model testing>
```

::::::::::::

::: instructor

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results/snps
module load bioinfo-tools iqtree
iqtree2 -s core.aln -m MFP+ASC
```

::::::::::::

:::::::::::::

@@

```bash
cd tree
mv core.aln.treefile core.aln.newick 
cat core.aln.newick
```

@@

What model was selected? How does IQ-TREE choose between different models? What does e.g. "JC+ASC+G4" means?

::: hint

The log file of IQ-TREE contains a lot of information.

The [IQ-TREE website](http://www.iqtree.org/doc/Substitution-Models#dna-models) contains a description of the DNA substitution models.

:::::::::

@@

IQ-TREE chose the TVMe+ASC model. We've discussed the ASC above, and you can read more about the TVMe in the [IQ-TREE manual](http://www.iqtree.org/doc/Substitution-Models#dna-models). In short, it is a mdoel where base frequencies are deemed equal, and all rates estimated independently, except the rates of transversions (A <-> G and C <-> T) are equal.

The model selection process is also described on the [IQ-TREE website](http://www.iqtree.org/doc/Tutorial#choosing-the-right-substitution-model).

@@

## Challenge: Visualize the tree on your computer

Copy the tree in Newick format (`core.aln.newick`) to your computer using `scp` and use FigTree or (phylo.io)[https://beta.phylo.io/] to visualize it. Do you see a way to assess the quality of the tree?

```bash
scp <username>@rackham.uppmax.uu.se:<file with path> .
```

::::::::::::

::: instructor

```bash
scp <username>@rackham.uppmax.uu.se:/proj/g2020004/nobackup/3MK013/<user>/molepi/results/tree/core.aln.newick .
```

::::::::::::

## Challenge: Rerun the tree, calculating bootstraps this time

In the [Phylogenetics](phylogenetics.Rmd) module, we calculated 1000 UltraFast bootstraps using the `-B 1000` option. Here, calculate 100 non-parametric bootstraps. [Non-parametric](https://en.wikipedia.org/wiki/Nonparametric_statistics) bootstraps are bootstraps that are calculated without making assumptions about the data. In that case, it means actually resampling the alignment and calculating trees for each of the bootstrapped alignments. It is a much slower process than the UltraFast method (which is parametric), but in that case, since the alignment is very small, the computational load is acceptable.

With IQ-TREE, [non-parametric bootstraps](http://www.iqtree.org/doc/Tutorial#assessing-branch-supports-with--standard-nonparametric-bootstrap) are inferred using the `-b <n bootstraps>` option. Make sure all files are prefixed with `core.aln.100npb` (option `--prefix`).

The process will take a minute or so. Then in the `results` folder, next to the `tree` folder, create a `tree100npb` folder and move the results file there.

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results/snps
iqtree2 <alignment> <model as before> <100 non-parametric bootstraps> <prefix>
cd ..
mkdir tree100npb
mv snps/core.aln.100npb* tree100npb/
```

::::::::::::

::: instructor

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results/snps
iqtree2 -s core.aln -m MFP+ASC -b 100 --prefix core.aln.100npb
cd ..
mkdir tree100npb
mv snps/core.aln.100npb* tree100npb/

```

::::::::::::

When that finishes, copy the resulting tree to your local computer and visualize it. Display support values:
- With FigTree: on the left menu, tick "Branch labels", and choose the last row (what you entered when prompted to do so, `label` by default)
- With phylo.io: do *not* use the "beta" service, as it displays the support values incorrectly. Use the stable version, but compare both!

@@

![As shown by FigTree](fig/figtree.png){alt='Figtree'}
![As shown by phylo.io, stable version](fig/phylo_io.png){alt='phylo.io'}
  ![As shown by phylo.io, beta version. Note the position of the "99"](fig/phylo_io_beta.png){alt='beta.phylo.io'}
